% !TEX root = ../main.tex

\chapter{Experimental Setup}\label{chapter:experiments}

This chapter describes the dataset, training protocol, and evaluation methodology underlying the eight training trials and four UQ experiments reported in this thesis.

\section{Dataset}

\subsection{Paris Road Network}

Our experiments use a MATSim simulation of the Paris metropolitan area, provided by Natterer~et~al.~\cite{natterer2025ml}. Key characteristics:

\begin{table}[htbp]
\centering
\caption[Paris road network dataset statistics]{Dataset statistics for the Paris MATSim road network.}
\label{tab:dataset}
\begin{tabular}{lr}
\toprule
\textbf{Property} & \textbf{Value} \\
\midrule
Road segments (nodes) & 31,635 \\
Network edges & $\sim$85,000 \\
Available scenarios (full dataset) & 10,000 \\
Scenarios used in this work & 1,000 (10\% subset) \\
Training scenarios (T1--T6, 80/15/5) & 800 \\
Training scenarios (T7--T8, 80/10/10) & 800 \\
Test scenarios (T1--T6) & 50 \\
Test scenarios (T7--T8) & 100 \\
Total test predictions (T1--T6) & 1,581,750 \\
Total test predictions (T7--T8) & 3,163,500 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Note on data scale:} All experiments in this thesis were conducted using only \textbf{1,000 of 10,000 available MATSim scenarios (10\% subset)}, unless stated otherwise. The full 10,000-scenario dataset was used in the original work by Natterer~et~al.~\cite{natterer2025ml}.

\subsection{Scenario Generation}

Each scenario represents a distinct policy intervention. A random subset of road segments is selected and assigned a capacity reduction ranging from 10\% to 100\%, with scenarios designed to cover a diverse range of intervention patterns across the network.

\subsection{Data Split}

Training trials T1--T6 use an \textbf{80/15/5} split (train/validation/test). Trials T7--T8 use an \textbf{80/10/10} split, providing a larger test set (100 graphs, 3,163,500 predictions) for more reliable UQ evaluation. The split is at the \emph{scenario level}---each scenario appears in exactly one split.

\section{Training Protocol}

\subsection{Model Training}

We train 8 models with varying hyperparameters. All trials use the PointNetTransfGAT architecture~\cite{natterer2025ml}. Trial~1 differs from Trials~2--8 in its final prediction layer: Trial~1 uses a \texttt{Linear(64$\to$1)} output layer, while Trials~2--8 use \texttt{GATConv(64$\to$1)}.

\begin{table}[htbp]
\centering
\caption[Training trials and hyperparameter configurations]{Training trials and configurations. T1 uses \texttt{Linear(64$\to$1)} final layer; T2--T8 use \texttt{GATConv(64$\to$1)}. Note the different data splits between trials.}
\label{tab:trials}
\begin{tabular}{lcccccc}
\toprule
\textbf{Trial} & \textbf{Batch} & \textbf{Split} & \textbf{Dropout} & \textbf{LR} & \textbf{Weighted} & \textbf{Final Layer} \\
\midrule
T1 & 32 & 80/15/5 & 0.0$^\dagger$ & $10^{-3}$ & No & Linear \\
T2 & 16 & 80/15/5 & 0.3 & $5{\times}10^{-4}$ & No & GATConv \\
T3 & 16 & 80/15/5 & 0.0 & $5{\times}10^{-4}$ & Yes & GATConv \\
T4 & 16 & 80/15/5 & 0.0 & $5{\times}10^{-4}$ & Yes & GATConv \\
T5 & 8 & 80/15/5 & 0.3 & $5{\times}10^{-4}$ & No & GATConv \\
T6 & 8 & 80/15/5 & 0.3 & $3{\times}10^{-4}$ & No & GATConv \\
T7 & 8 & 80/10/10 & 0.3 & $6{\times}10^{-4}$ & No & GATConv \\
\textbf{T8} & \textbf{8} & \textbf{80/10/10} & \textbf{0.2} & $\mathbf{5{\times}10^{-4}}$ & \textbf{No} & \textbf{GATConv} \\
\bottomrule
\end{tabular}
\vspace{2pt}
\begin{minipage}{\linewidth}
\footnotesize $^\dagger$ T1 was instantiated with architecture parameter \texttt{dropout=0.3}, but \texttt{use\_dropout=False} disabled dropout during both training and inference; the effective dropout rate is 0.0. Source: \texttt{MODEL\_SUMMARY.md} for Trial~1.
\end{minipage}
\end{table}

\subsection{Optimization}

All trials use the Adam optimizer~\cite{kingma2015adam} with mean squared error (MSE) loss, or weighted MSE where noted. Batch size varies by trial (8--32 scenarios) and learning rates are given in Table~\ref{tab:trials}. Training was conducted on an NVIDIA T4 GPU via Google Colab.

\subsection{Framework}

All models are implemented in PyTorch 2.0~\cite{paszke2019pytorch} using PyTorch Geometric~\cite{fey2019fast} for graph operations, with Weights \& Biases for experiment tracking.

\section{UQ Experimental Design}

\subsection{MC Dropout Evaluation (T5--T8)}

MC Dropout uncertainty ($S = 30$ stochastic forward passes) was evaluated on Trials 5--8, which include non-zero dropout rates. The key metric is Spearman~$\rho$ between predicted uncertainty ($\sigma$) and absolute prediction error~($|y - \hat{y}|$). Trials~5 and~6 were evaluated on 50 test graphs (1,581,750 nodes), while Trials~7 and~8 used the larger 100-graph test set (3,163,500 nodes). T8 (dropout $= 0.2$, batch $= 8$) was selected as the primary UQ model based on best test $R^2$ among comparable trials (T2--T8).

\subsection{Experiment A: MC Dropout vs Ensemble Variance}

Experiment~A compares MC Dropout and ensemble variance as uncertainty estimators using the same base model, T8. MC Dropout is run with $S = 30$ samples and compared against the variance across 5 independently seeded training runs, with both methods evaluated on the full 100-scenario test set (3,163,500 predictions).

One important caveat: the ensemble runs in Experiment~A were evaluated on a subset that exhibited a data distribution mismatch (near-zero $R^2 \approx 0.003$). This is a known artifact of the ensemble evaluation setup and does not reflect individual model quality. Ensemble Spearman~$\rho$ values should be interpreted in this context.

\subsection{Experiment B: Multi-Model Ensemble}

Experiment~B asks whether combining multiple trained models improves uncertainty quantification. Five trained models (T2, T5, T6, T7, T8) are assembled into a multi-model ensemble; their disagreement on each prediction serves as the uncertainty estimate, and ensemble Spearman~$\rho$ is compared against the single-model MC Dropout result.

As with Experiment~A, Experiment~B exhibits the same data distribution mismatch. Results are reported for completeness but should not be compared directly to single-model MC Dropout metrics.

\subsection{Conformal Prediction (T8)}

Split conformal prediction~\cite{angelopoulos2023conformal} is applied to T8 using the 100-graph test set, divided equally into 50 calibration graphs and 50 evaluation graphs. The nonconformity score is the absolute residual $|y - \hat{y}|$, and prediction intervals are constructed at both 90\% and 95\% coverage levels.

\section{Computational Resources}

All experiments were run on Google Colab Pro with an NVIDIA T4 GPU (16\,GB VRAM). Total training time across the eight trials was approximately 13 hours. MC Dropout inference for T8 ($S = 30$, 100 test graphs) required approximately 228 minutes.
